Reconstructing the last known movements of one of Nature's giants

### 2017-11-15 manuscript/revision/blue-whale-ms-v2.tex
Few edits to figure legend

% Edit figure legend please

% figure 1
%\begin{figure}
 % \centering
  %\includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  %\caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes as well as a change in the mean and magnitude of the cycles during 1889 (approximately at sample 30cm). Approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  %\label{fig1}
%\end{figure}

% Krill has a very similar isotopic composition to phytoplankton

@@

%line 112 - 122 para
The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

@@

Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}. 
However, a growing body of evidence suggests that not all whales migrate to subtropical waters each year, with some remaining at high latitudes year round\cite{mcdonald2006biogeographic}. %line 127
Our results support this conclusion. 
Interestingly, there is still a resident%line 128
 population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
Whaling was an intense pressure for blue whales during the period we are analysing. 
Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
The NHM blue whale thus represents an individual from a species at the brink of extinction.
Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures. % line 137 and 139

>>>>

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{revision/figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

@@

The last 500 days of the whale's life are difficult to simulate. 
Beginning in the winter of 1889/1890, the observed $\delta^{13}$C profile is relatively depleted, and remains constant for approximately six months, before increasing rapidly in the winter of 1890/1891. 
Depleted $\delta^{13}$C values are found in northern waters that show large temporal fluctuations in $\delta^{13}$Cplk values [fig SX residency plot], and the observed values cannot be simulated purely from movements within the known geographic range.
We infer that the $\delta^{13}$C record in the last period of life reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions\cite{cardona2017temporal}. 
We suggest that the constant, depleted $\delta^{13}$C values reflect use of carbon reserves assimilated in northern latitudes during 6-8 months residency in low latitude waters. 
After six months the whale began feeding in low latitude waters producing a dramatic increase in $\delta^{13}$C values and began a final migration north. 
Blue whales have a 10-12 month gestation period, with calving occurring in subtropical waters, and calves are weaned after 6-7 months\cite{handbook}. 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, residency in warm waters through the whole of 1890 with feeding resuming in the second half of 1890 and eventual stranding during the return to northern feeding grounds in early 1891, is therefore consistent with the known reproductive ecology of blue whales.

@@

Conventional wisdom suggests that baleen whales migrate each year from feeding grounds near the poles in summer, to overwintering and breeding grounds nearer the equator\cite{corkeron1999baleen,lockyer1981migration}. 
However, a growing body of evidence suggests that not all whales migrate each year, with some remaining at high or low latitudes year round\cite{mcdonald2006biogeographic,busquets2017estimating}. 
Our results support this conclusion. 
Interestingly, there is still a resident%line 128
 population of blue whales in the regions indicated by our model, and females still breed in the subtropical waters near the Azores\cite{reilly2008balaenoptera}. 
Whaling was an intense pressure for blue whales during the period we are analysing. 
Before whaling in the North Atlantic began in 1868\cite{reilly2008balaenoptera}, there were 10,000-15,000 blue whales in the region\cite{sigurjonsson1995life}. 
In the early 20th century, fisheries moved outside the area because stocks were so depleted\cite{reilly2008balaenoptera}; during this period over 12,000 blue whales were landed\cite{sigurjonsson1995life}. 
The NHM blue whale thus represents an individual from a species at the brink of extinction.
Our data suggest that the spatial ecology of blue whales in the North Atlantic has remained similar for more than 100 years, despite dramatic changes in anthropogenic pressures. % line 137 and 139
